---
title: "Scratch Pad"
author: "Bryan Johns"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(sf)
library(ggspatial)
```

# Scratch Pad for 2024 Cyclistic Data Analysis

Exploratory Data Analysis, unencumbered by the need to be **too** presentable. 

## Load data

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# load data, combine into a df
read_cyclistic_data <- function(year) {
  for (i in 1:12) {
    month <- sprintf("%02d", i)  # Format as 01, 02, ..., 12
    file_path <- paste0("resources/data/", year, "/", year, month, "-divvy-tripdata_cleaned.csv")
    
    if (file.exists(file_path)) {
      data <- read_csv(file_path)
      data$month <- as.character(data$month)  # ensure month is character
      assign(paste0("m", month), data, envir = .GlobalEnv)
    } else {
      message("File not found: ", file_path)
    }
  }
}

read_cyclistic_data(2024)

# regex to bind rows of all monthly data frames
df <- bind_rows(mget(ls(pattern = "^m\\d{2}$")))

# dump memory
rm(list = ls(pattern = "^m\\d{2}(_cleaned)?$"))
gc()
```

Number of member vs casual rides

```{r}
table(df$member_casual)
```

3.7 million member rides vs 2.2 million casual rides. Unfortunately, there is no way to tell how many are repeat users. Presumably, many member riders repeat far more often than casual riders. The scenario puts members at 30% of Cyclistic.

In the scenario, and in real life, the data is shared without any sort of identifying information that would help track that, probably why after 2019 they removed gender and birth year. It wasn't much help anyway.

```{r}
table(df$month)
```

Reassigning the month "9" to "09".

```{r}
df$month <- sprintf("%02d", as.numeric(format(df$date, "%m")))
table(df$month)
```


## NA stations and e-bikes

```{r}
# NA value counts
na_counts <- sapply(df, function(x) sum(is.na(x)))

# print
print(na_counts)
```

Start station counts by rideable_type

```{r}
df %>%
  mutate(missing_name = is.na(start_station_name)) %>%
  group_by(rideable_type) %>%
  summarize(missing_count = sum(missing_name), total = n(), perc_missing = mean(missing_name)) %>%
  arrange(desc(perc_missing))
```

Yup. Missing stations are electric bikes without a docking station, left randomly throughout the city.

End stations counts by rideable_type

```{r}
df %>%
  mutate(missing_name = is.na(end_station_name)) %>%
  group_by(rideable_type) %>%
  summarize(missing_count = sum(missing_name), total = n(), perc_missing = mean(missing_name)) %>%
  arrange(desc(perc_missing))

```

```{r}
known_stations <- df %>%
  filter(!is.na(start_station_id) & !is.na(start_station_name)) %>%
  select(start_station_id, start_station_name, start_lat, start_lng) %>%
  distinct()
print(count(known_stations))
```

```{r}
# attempt to consolidate known stations
known_stations <- df %>%
  filter(!is.na(start_station_id) & !is.na(start_station_name)) %>%
  mutate(
    start_lat = round(start_lat, 4), # roughly the length of a bus
    start_lng = round(start_lng, 4)
  ) %>%
  select(start_station_id, start_station_name, start_lat, start_lng) %>%
  distinct()
print(count(known_stations))
```

Stations no longer have a specific lat/lng. No exact station location maps anymore. Have to do heatmaps.

```{r}
# distinct start_station_name count
distinct_start_station_count <- df %>%
  summarise(distinct_count = n_distinct(start_station_name))

# print
print(distinct_start_station_count)
```
 It's supposed to be 692, per the scenario. (I found 608 in 2019).

```{r}
# distinct start_station_id count
distinct_start_station_count <- df %>%
  summarise(distinct_count = n_distinct(start_station_id))

# print
print(distinct_start_station_count)
```

So, 45 duplicate station names, probably easily explained. Easier than the extra ~1000 station id's.

Electronic bikes are a game changer. I thought switching from 2019 to 2024 was gonna be easy.

## Ride length (and NA end_lat/lng)

IQR

```{r}
summary(df$ride_length)
```

The mean is higher than the 3rd quartile. Data set skews left. Makes sense, as many commuter rides are quick, but leisure rides can last a while. And sometimes people just dump the bike in a ditch so it appears to be a long ride.

```{r}
max(df$ride_length)/60/24
```

A little over a day. Somebody forgot to dock a bike.

```{r}
print(df[which.max(df$ride_length), ])
```

Average of rides with no end lat(/lng also missing)

```{r}
# average ride_length for rows with missing end_lat
avg_ride_length_missing_end_lat <- df %>%
  filter(is.na(end_lat)) %>%
  summarise(avg_ride_length = mean(ride_length, na.rm = TRUE))

# print
print(avg_ride_length_missing_end_lat)/60/24
```

A little over a day.

Min of rides with no end lat(/lng also missing)

```{r}
# minimum ride_length for rows with missing end_lat
min_ride_length_missing_end_lat <- df %>%
  filter(is.na(end_lat)) %>%
  summarise(min_ride_length = min(ride_length, na.rm = TRUE))

# print
print(min_ride_length_missing_end_lat)/60/24
```

```{r}
## histogram of ride_length
ggplot(df, aes(x = ride_length)) +
  geom_histogram(binwidth = 5, fill = "blue", color = "black", alpha = 0.7) +
  scale_x_continuous(limits = c(0, quantile(df$ride_length, 0.99))) + # trim extreme outliers
  labs(title = "Distribution of Ride Length",
       subtitle = "Removed the 99th percentile outliers - the 1%!",
       x = "Ride Length (minutes)",
       y = "Count") +
  theme_minimal()
```

## Ride length outliers / IQR

Note that I can toggle between these and "run all" to see the impact... and the power to skew results. Mwahahahaha!

```{r}
# IQR * 1.5 for outliers
# calculate Q1, Q3, and IQR
Q1 <- quantile(df$ride_length, 0.25, na.rm = TRUE)
Q3 <- quantile(df$ride_length, 0.75, na.rm = TRUE)
IQR <- Q3 - Q1

# lower and upper bounds for outliers
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

# filter out outliers
df_no_outliers <- df %>%
  filter(ride_length >= lower_bound & ride_length <= upper_bound)

# print
summary(df_no_outliers$ride_length)
```

```{r}
# IQR * 3 for outliers
lower_bound_strict <- Q1 - 3 * IQR
upper_bound_strict <- Q3 + 3 * IQR

# filter out outliers
df_no_outliers <- df %>%
  filter(ride_length >= lower_bound_strict & ride_length <= upper_bound_strict)

# print
summary(df_no_outliers$ride_length)
```

```{r}
# remove the 99th percentile
upper_1_percent <- quantile(df$ride_length, 0.99, na.rm = TRUE)

# filter out the top 1% of the data
df_no_outliers <- df %>%
  filter(ride_length <= upper_1_percent)

# print
summary(df_no_outliers$ride_length)
```

```{r}
# the 99.5 percentile
upper_.5_percent <- quantile(df$ride_length, 0.995, na.rm = TRUE)

# filter out the top 0.5% of the data
df_no_outliers <- df %>%
  filter(ride_length <= upper_.5_percent)

# print
summary(df_no_outliers$ride_length)
```

```{r}
# the 99.75 percentile
upper_.25_percent <- quantile(df$ride_length, 0.9975, na.rm = TRUE)

# filter out the top 0.025% of the data
df_no_outliers <- df %>%
  filter(ride_length <= upper_.25_percent)

# print
summary(df_no_outliers$ride_length)


```

I will very conservatively drop the 0.25% of rides over four and a half hours long.

# Visualizations

## Rides by Member type (total rides, duration)

```{r}
# sum total ride length by membership type
total_minutes_ridden <- df %>%
  group_by(member_casual) %>%
  summarise(total_minutes = sum(ride_length, na.rm = TRUE))

# bar chart of total minutes ridden
ggplot(total_minutes_ridden, aes(x = member_casual, y = total_minutes, fill = member_casual)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = round(total_minutes)), vjust = -0.5) +
  theme_minimal() +
  labs(title = "Total Minutes Ridden by Membership Type", y = "Total Minutes Ridden", x = "Rider Type")

```

```{r}
# sum total ride length by membership type
total_minutes_ridden <- df_no_outliers %>%
  group_by(member_casual) %>%
  summarise(total_minutes = sum(ride_length, na.rm = TRUE))

# bar chart of total minutes ridden
ggplot(total_minutes_ridden, aes(x = member_casual, y = total_minutes, fill = member_casual)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = round(total_minutes)), vjust = -0.5) +
  theme_minimal() +
  labs(title = "Total Minutes Ridden by Membership Type, No Outliers", y = "Total Minutes Ridden", x = "Rider Type")

```

```{r}
# sum average ride length by membership type
average_ride_length <- df %>%
  group_by(member_casual) %>%
  summarise(average_minutes = mean(ride_length, na.rm = TRUE))

# bar chart of average ride length
ggplot(average_ride_length, aes(x = member_casual, y = average_minutes, fill = member_casual)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = round(average_minutes, 1)), vjust = -0.5) +
  theme_minimal() +
  labs(title = "Average Ride Length by Membership Type", y = "Average Ride Length (minutes)", x = "Rider Type")
```

```{r}
# sum average ride length by membership type
average_ride_length <- df_no_outliers %>%
  group_by(member_casual) %>%
  summarise(average_minutes = mean(ride_length, na.rm = TRUE))

# bar chart of average ride length
ggplot(average_ride_length, aes(x = member_casual, y = average_minutes, fill = member_casual)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = round(average_minutes, 1)), vjust = -0.5) +
  theme_minimal() +
  labs(title = "Average Ride Length by Membership Type, No Outliers", y = "Average Ride Length (minutes)", x = "Rider Type")
```

```{r}
# bar chart 0f ride frequency per user
ggplot(df, aes(x = member_casual, fill = member_casual)) +
  geom_bar() +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
  labs(title = "Total Rides by Membership Type",
       x = "Rider Type",
       y = "Total Rides") +
  theme_minimal()

```

```{r}
# bar chart 0f ride frequency per user
ggplot(df_no_outliers, aes(x = member_casual, fill = member_casual)) +
  geom_bar() +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
  labs(title = "Total Rides by Membership Type, No Outliers",
       x = "Rider Type",
       y = "Total Rides") +
  theme_minimal()

```

```{r}
# histograms of ride length per user type
ggplot(df, aes(x = ride_length, fill = member_casual)) +
  geom_histogram(binwidth = 5, alpha = 0.7) +
  facet_wrap(~ member_casual, scales = "free_y") +
  labs(title = "Histogram of Ride Lengths by Membership Type",
       x = "Ride Length (minutes)",
       y = "Frequency") +
  theme_minimal() +
  # scale_fill_manual(values = c("member" = "blue", "casual" = "red")) +
  xlim(0,100)
```

Members ride more frequently for shorter duration (probably commuting).

Casuals ride longer (probably leisure).

## Time and Ridership

```{r}
# line chart of rides per month
df %>%
  group_by(month, member_casual) %>%
  summarise(rides = n()) %>%
  ggplot(aes(x = month, y = rides, color = member_casual, group = member_casual)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(title = "Rides per Month, by Customer Type", x = "Month", y = "Number of Rides")
```

Peak in summer, unsurprisingly.

```{r}
# bar chart of rides per month
df %>%
  group_by(month, member_casual) %>%
  summarise(rides = n()) %>%
  ggplot(aes(x = month, y = rides, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = rides), vjust = -0.5, position = position_dodge(0.9)) +
  theme_minimal() +
  labs(title = "Total Rides per Month by Customer Type", x = "Month", y = "Number of Rides")
```

```{r}
# bar chart of ride length per month
df %>%
  group_by(month, member_casual) %>%
  summarise(avg_ride_length = mean(ride_length, na.rm = TRUE)) %>%
  ggplot(aes(x = month, y = avg_ride_length, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = round(avg_ride_length, 1)), vjust = -0.5, position = position_dodge(0.9)) +
  theme_minimal() +
  labs(title = "Average Trip Duration per Month by Customer Type", x = "Month", y = "Average Trip Duration (minutes)")
```

Here is where I noticed a "9" month. There were 35 rides in the month "9" instead of "09". Stuck out at 100 and 450 minute long trip durations (outliers). I fixed it up at the top.


```{r}
# rides per season
df %>%
  mutate(season = factor(season, levels = c("Winter", "Spring", "Summer", "Fall"))) %>%
  group_by(season, member_casual) %>%
  summarise(rides = n()) %>%
  ggplot(aes(x = season, y = rides, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = rides), vjust = -0.5, position = position_dodge(0.9)) +
  theme_minimal() +
  labs(title = "Total Rides per Season by Customer Type", x = "Season", y = "Number of Rides") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2)))
```

```{r}
# ride length per season
df %>%
  mutate(season = factor(season, levels = c("Winter", "Spring", "Summer", "Fall"))) %>%
  group_by(season, member_casual) %>%
  summarise(avg_ride_length = mean(ride_length, na.rm = TRUE)) %>%
  ggplot(aes(x = season, y = avg_ride_length, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = round(avg_ride_length, 1)), vjust = -0.5, position = position_dodge(0.9)) +
  theme_minimal() +
  labs(title = "Average Trip Duration per Season by Customer Type", x = "Season", y = "Average Trip Duration (minutes)") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2)))
```


```{r}
# line chart of rides per hour
df %>%
  group_by(hour, member_casual) %>%
  summarise(rides = n()) %>%
  ggplot(aes(x = hour, y = rides, color = member_casual, group = member_casual)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(title = "Rides per Hour, by Customer Type", x = "Hour", y = "Number of Rides")
```

```{r}
# bar chart of rides per hour
df %>%
  group_by(hour, member_casual) %>%
  summarise(rides = n()) %>%
  ggplot(aes(x = hour, y = rides, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(title = "Rides per Hour of Day by Customer Type", x = "Hour of Day", y = "Number of Rides")
```

```{r}
# bar chart of rides per hour by day of week, free y-axis
df %>%
  mutate(day_of_week = factor(day_of_week, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))) %>%
  group_by(hour, member_casual, day_of_week) %>%
  summarise(rides = n()) %>%
  ggplot(aes(x = hour, y = rides, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~ day_of_week, scales = "free_y") +
  theme_minimal() +
  labs(title = "Rides per Hour of Day by Customer Type and Day of Week", subtitle = "Free y-axis", x = "Hour of Day", y = "Number of Rides")
```

```{r}
# bar chart of rides per hour by day of week, locked y-axis
df %>%
  mutate(day_of_week = factor(day_of_week, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))) %>%
  group_by(hour, member_casual, day_of_week) %>%
  summarise(rides = n()) %>%
  ggplot(aes(x = hour, y = rides, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~ day_of_week) +
  theme_minimal() +
  labs(title = "Rides per Hour of Day by Customer Type and Day of Week", subtitle = "Locked y-axis", x = "Hour of Day", y = "Number of Rides")
```

```{r}
# bar chart of rides / member_casual / day of week
df %>%
  mutate(day_of_week = factor(day_of_week, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))) %>%
  group_by(member_casual, day_of_week) %>%
  summarise(rides = n()) %>%
  ggplot(aes(x = member_casual, y = rides, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = rides), vjust = -0.5, position = position_dodge(0.9)) +
  facet_wrap(~ day_of_week) +
  theme_minimal() +
  labs(title = "Rides per Member Type by Day of Week", x = "Member Type", y = "Number of Rides") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2)))
```

```{r}
# bar chart of rides / member_casual / day of week
df %>%
  mutate(day_of_week = factor(day_of_week, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))) %>%
  group_by(member_casual, day_of_week) %>%
  summarise(rides = n()) %>%
  ggplot(aes(x = day_of_week, y = rides, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = rides), vjust = -0.5, position = position_dodge(0.9)) +
  theme_minimal() +
  labs(title = "Rides per Member Type by Day of Week", x = "Member Type", y = "Number of Rides") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2)))
```

```{r}
# bar chart avg ride time / member_casual / day of week
df %>%
  mutate(day_of_week = factor(day_of_week, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))) %>%
  group_by(member_casual, day_of_week) %>%
  summarise(avg_ride_length = mean(ride_length, na.rm = TRUE)) %>%
  ggplot(aes(x = member_casual, y = avg_ride_length, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = round(avg_ride_length, 1)), vjust = -0.5, position = position_dodge(0.9)) +
  facet_wrap(~ day_of_week) +
  theme_minimal() +
  labs(title = "Average Ride Length per Member Type by Day of Week", x = "Member Type", y = "Average Ride Length (minutes)") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2)))
```

```{r}
# bar chart avg ride time / member_casual / day of week
df_no_outliers %>%
  mutate(day_of_week = factor(day_of_week, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))) %>%
  group_by(member_casual, day_of_week) %>%
  summarise(avg_ride_length = mean(ride_length, na.rm = TRUE)) %>%
  ggplot(aes(x = member_casual, y = avg_ride_length, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = round(avg_ride_length, 1)), vjust = -0.5, position = position_dodge(0.9)) +
  facet_wrap(~ day_of_week) +
  theme_minimal() +
  labs(title = "Average Ride Length per Member Type by Day of Week, No Outliers", x = "Member Type", y = "Average Ride Length (minutes)") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2)))
```

```{r}
# bar chart avg ride time / member_casual / day of week
df %>%
  mutate(day_of_week = factor(day_of_week, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))) %>%
  group_by(member_casual, day_of_week) %>%
  summarise(avg_ride_length = mean(ride_length, na.rm = TRUE)) %>%
  ggplot(aes(x = day_of_week, y = avg_ride_length, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = round(avg_ride_length, 1)), vjust = -0.5, position = position_dodge(0.9)) +
  theme_minimal() +
  labs(title = "Average Ride Length per Member Type by Day of Week", x = "Member Type", y = "Average Ride Length (minutes)") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2)))
```

```{r}
# bar chart avg ride time / member_casual / day of week
df_no_outliers %>%
  mutate(day_of_week = factor(day_of_week, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))) %>%
  group_by(member_casual, day_of_week) %>%
  summarise(avg_ride_length = mean(ride_length, na.rm = TRUE)) %>%
  ggplot(aes(x = day_of_week, y = avg_ride_length, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = round(avg_ride_length, 1)), vjust = -0.5, position = position_dodge(0.9)) +
  theme_minimal() +
  labs(title = "Average Ride Length per Member Type by Day of Week, No Outliers", x = "Member Type", y = "Average Ride Length (minutes)") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2)))
```

```{r}
# aggregate member rider counts by hour and day of the week
member_peak_times <- df %>%
  mutate(day_of_week = factor(day_of_week, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))) %>%  filter(member_casual == "member") %>%
  group_by(day_of_week, hour) %>%
  summarise(rides = n(), .groups = "drop")

# heatmap of peak times for member riders
ggplot(member_peak_times, aes(x = hour, y = day_of_week, fill = rides)) +
  geom_tile() +
  scale_fill_viridis_c() +
  theme_minimal() +
  labs(title = "Peak Usage Times for Member Riders", x = "Hour of Day", y = "Day of Week")
```

Members peak with commuter hours.

```{r}
# aggregate casual rider counts by hour and day of the week
casual_peak_times <- df %>%
  mutate(day_of_week = factor(day_of_week, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))) %>%  filter(member_casual == "casual") %>%
  group_by(day_of_week, hour) %>%
  summarise(rides = n(), .groups = "drop")

# heatmap of peak times for casual riders
ggplot(casual_peak_times, aes(x = hour, y = day_of_week, fill = rides)) +
  geom_tile() +
  scale_fill_viridis_c() +
  theme_minimal() +
  labs(title = "Peak Usage Times for Casual Riders", x = "Hour of Day", y = "Day of Week")
```

Peak time is Saturday afternoon, followed by Sunday afternoon. But also a little around peak end-of-work hour. Are casual riders transiting in to work, and biking back?

## Bike Type

```{r}
# bar chart of total count of rideable_type
ggplot(df, aes(x = rideable_type, fill = rideable_type)) +
  geom_bar() +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
  theme_minimal() +
  labs(title = "Total Count of Rideable Types",
       x = "Rideable Type",
       y = "Count")
```

```{r}
# bar chart of rideable_type count per member_casual
ggplot(df, aes(x = rideable_type, fill = member_casual)) +
  geom_bar(position = "dodge") +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5, position = position_dodge(0.9)) +
  theme_minimal() +
  labs(title = "Count of Rideable Types per Member Type",
       x = "Rideable Type",
       y = "Count")
```


## Stations

```{r}
# bar chart: top start stations
df %>%
  count(start_station_name, member_casual, sort = TRUE) %>%
  group_by(member_casual) %>%
  top_n(20, n) %>%
  ggplot(aes(x = reorder(start_station_name, n), y = n, fill = member_casual)) +
  geom_col() +
  coord_flip() +
  theme_minimal() +
  labs(title = "Top Start Stations by Rider Type", x = "Station Name", y = "Number of Rides")
```

```{r}
# bar chart: top start stations
df %>%
  filter(!is.na(start_station_name)) %>% 
  count(start_station_name, member_casual, sort = TRUE) %>%
  group_by(member_casual) %>%
  top_n(20, n) %>%
  ggplot(aes(x = reorder(start_station_name, n), y = n, fill = member_casual)) +
  geom_col() +
  coord_flip() +
  theme_minimal() +
  labs(title = "Top Start Stations by Rider Type", x = "Station Name", y = "Number of Rides")
```

"Milleniumn Park", "Shedd Aquarium", "Theater on the Lake", "Dusable Lake Shore / Harbor", and "Adler Planetarium" sound like destinations for the occasional bike rider.

```{r}
# bar chart: top end stations
df %>%
  filter(!is.na(end_station_name)) %>% 
  count(end_station_name, member_casual, sort = TRUE) %>%
  group_by(member_casual) %>%
  top_n(20, n) %>%
  ggplot(aes(x = reorder(end_station_name, n), y = n, fill = member_casual)) +
  geom_col() +
  coord_flip() +
  theme_minimal() +
  labs(title = "Top End Stations by Rider Type", x = "Station Name", y = "Number of Rides")
```

```{r}
# most popular start stations
top_stations <- df %>%
  filter(!is.na(start_station_name)) %>% 
  count(start_station_name, member_casual, sort = TRUE) %>%
  group_by(member_casual) %>%
  top_n(20, n)

print(top_stations)
```

```{r}
# most popular end stations
top_stations <- df %>%
  filter(!is.na(end_station_name)) %>% 
  count(end_station_name, member_casual, sort = TRUE) %>%
  group_by(member_casual) %>%
  top_n(20, n)

print(top_stations)
```

```{r}
# Leaflet: popular start stations with cleaned data
library(leaflet)

# remove rows with missing latitude or longitude values
cleaned_data <- df %>%
  filter(!is.na(start_lng) & !is.na(start_lat))

# get a sample - too many rows, it crashes my computer to plot all 4 million
set.seed(123)  # for reproducibility of sample
cleaned_data <- cleaned_data %>% sample_n(1000)

# plot sampled stations
leaflet(data = cleaned_data) %>%
  addTiles() %>%
  addMarkers(lng = ~start_lng, lat = ~start_lat, clusterOptions = markerClusterOptions())
```

Remember, there are 1800 stations (per the data), not 692 (per the company - in scenario), for mysterious reasons.

```{r}
# Leaflet: popular end stations with cleaned data
library(leaflet)

# remove rows with missing latitude or longitude values
cleaned_data <- df %>%
  filter(!is.na(end_lng) & !is.na(end_lat))

# get a sample - too many rows, it crashes my computer to plot all 4 million
set.seed(123)  # for reproducibility of sample
cleaned_data <- cleaned_data %>% sample_n(1000)

# plot sampled stations
leaflet(data = cleaned_data) %>%
  addTiles() %>%
  addMarkers(lng = ~end_lng, lat = ~end_lat, clusterOptions = markerClusterOptions())
```

```{r}
library(sf)
library(ggspatial)

# sample data
set.seed(123)  # for reproducibility
cleaned_data <- df %>% sample_n(1000)

# create linestring geometries for routes
routes <- cleaned_data %>%
  drop_na(start_lng, start_lat, end_lng, end_lat) %>%
  rowwise() %>%
  mutate(geometry = st_sfc(st_linestring(matrix(c(start_lng, start_lat, end_lng, end_lat), ncol = 2, byrow = TRUE)), crs = 4326)) %>%
  st_as_sf()

# plot routes
ggplot() +
  annotation_map_tile(type = "osm", zoom = 12) +
  geom_sf(data = routes, aes(geometry = geometry), color = "blue", size = 0.5, alpha = 0.5) +
  labs(title = "Sample Routes",
       x = "Longitude",
       y = "Latitude") +
  theme_minimal()

```

```{r}
# aggregate casual rider counts by station
casual_peak_stations <- df %>%
  filter(!is.na(start_station_name)) %>% 
  filter(member_casual == "casual") %>%
  count(start_station_name, sort = TRUE) %>%
  top_n(10, n)

# bar chart of top stations for casual riders
ggplot(casual_peak_stations, aes(x = reorder(start_station_name, n), y = n, fill = start_station_name)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  theme_minimal() +
  labs(title = "Top Start Stations for Casual Riders", x = "Station Name", y = "Number of Rides")
```

```{r}
# map of top start stations for casual riders
library(sf)
library(ggspatial)

# convert top casual start stations to sf object
casual_start_sf <- casual_peak_stations %>%
  left_join(df, by = c("start_station_name")) %>% # Join to get lat/lng
  distinct(start_station_name, start_lat, start_lng) %>% # Remove duplicates
  drop_na(start_lat, start_lng) %>% 
  st_as_sf(coords = c("start_lng", "start_lat"), crs = 4326)

# plot the map
ggplot() +
  annotation_map_tile(type = "osm", zoom = 12) +
  geom_sf(data = casual_start_sf, color = "blue", size = 3, alpha = 0.7) +
  labs(title = "Top Start Stations for Casual Riders (Map)",
       x = "Longitude", y = "Latitude") +
  theme_minimal()

```

```{r}
# extract end stations for trips originating at top casual start stations
casual_end_from_start_sf <- df %>%
  filter(!is.na(end_station_name)) %>% 
  filter(start_station_name %in% casual_peak_stations$start_station_name) %>%
  distinct(start_station_name, end_station_name, end_lat, end_lng) %>%
  drop_na(end_lat, end_lng) %>%
  st_as_sf(coords = c("end_lng", "end_lat"), crs = 4326)

# map end stations
ggplot() +
  annotation_map_tile(type = "osm", zoom = 12) +
  geom_sf(data = casual_end_from_start_sf, color = "blue", size = 3, alpha = 0.7) +
  labs(title = "End Stations for Trips from Top Start Stations (Casual Riders)",
       x = "Longitude", y = "Latitude") +
  theme_minimal()

```


```{r}
# aggregate casual rider counts by station
casual_peak_stations <- df %>%
  filter(!is.na(end_station_name)) %>% 
  filter(member_casual == "casual") %>%
  count(end_station_name, sort = TRUE) %>%
  top_n(10, n)

# bar chart of top stations for casual riders
ggplot(casual_peak_stations, aes(x = reorder(end_station_name, n), y = n, fill = end_station_name)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  theme_minimal() +
  labs(title = "Top End Stations for Casual Riders", x = "Station Name", y = "Number of Rides")
```

```{r}
# Convert top casual end stations to sf object
casual_end_sf <- casual_peak_stations %>%
  filter(!is.na(end_station_name)) %>% 
  left_join(df, by = c("end_station_name")) %>% 
  distinct(end_station_name, end_lat, end_lng) %>% 
  drop_na(end_lat, end_lng) %>% 
  st_as_sf(coords = c("end_lng", "end_lat"), crs = 4326)

# Plot map
ggplot() +
  annotation_map_tile(type = "osm", zoom = 12) +
  geom_sf(data = casual_end_sf, color = "blue", size = 3, alpha = 0.7) +
  labs(title = "Top End Stations for Casual Riders (Map)",
       x = "Longitude", y = "Latitude") +
  theme_minimal()

```

Roughly looks like 2019.

```{r}
# extract start stations for trips ending at top casual end stations
casual_start_from_end_sf <- df %>%
  filter(!is.na(start_station_name)) %>% 
  filter(end_station_name %in% casual_peak_stations$end_station_name) %>%
  distinct(end_station_name, start_station_name, start_lat, start_lng) %>%
  drop_na(start_lat, start_lng) %>%
  st_as_sf(coords = c("start_lng", "start_lat"), crs = 4326)

# map: start stations
ggplot() +
  annotation_map_tile(type = "osm", zoom = 12) +
  geom_sf(data = casual_start_from_end_sf, color = "blue", size = 3, alpha = 0.7) +
  labs(title = "Start Stations for Trips Ending at Top End Stations (Casual Riders)",
       x = "Longitude", y = "Latitude") +
  theme_minimal()

```


```{r}
# aggregate member rider counts by station
member_peak_stations <- df %>%
  filter(!is.na(start_station_name)) %>% 
  filter(member_casual == "member") %>%
  count(start_station_name, sort = TRUE) %>%
  top_n(10, n)

# bar chart of top stations for member riders
ggplot(member_peak_stations, aes(x = reorder(start_station_name, n), y = n, fill = start_station_name)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  theme_minimal() +
  labs(title = "Top Start Stations for Member Riders", x = "Station Name", y = "Number of Rides")
```

```{r}
# convert top member start stations to sf object
member_start_sf <- member_peak_stations %>%
  left_join(df, by = c("start_station_name")) %>% 
  distinct(start_station_name, start_lat, start_lng) %>% 
  drop_na(start_lat, start_lng) %>% 
  st_as_sf(coords = c("start_lng", "start_lat"), crs = 4326)

# plot map
ggplot() +
  annotation_map_tile(type = "osm", zoom = 12) +
  geom_sf(data = member_start_sf, color = "red", size = 3, alpha = 0.7) +
  labs(title = "Top Start Stations for Member Riders (Map)",
       x = "Longitude", y = "Latitude") +
  theme_minimal()

```

```{r}
# extract end stations for trips starting at top member start stations
member_end_sf <- df %>%
  filter(!is.na(end_station_name)) %>% 
  filter(start_station_name %in% member_peak_stations$start_station_name) %>%
  distinct(start_station_name, end_station_name, end_lat, end_lng) %>%
  drop_na(end_lat, end_lng) %>%
  st_as_sf(coords = c("end_lng", "end_lat"), crs = 4326)

# plot map
ggplot() +
  annotation_map_tile(type = "osm", zoom = 12) +
  geom_sf(data = member_end_sf, color = "blue", size = 3, alpha = 0.7) +
  labs(title = "End Stations for Trips from Top Start Stations",
       x = "Longitude", y = "Latitude") +
  theme_minimal()

```

```{r}
# aggregate member rider counts by station
member_peak_stations <- df %>%
  filter(!is.na(end_station_name)) %>% 
  filter(member_casual == "member") %>%
  count(end_station_name, sort = TRUE) %>%
  top_n(10, n)

# bar chart of top stations for member riders
ggplot(member_peak_stations, aes(x = reorder(end_station_name, n), y = n, fill = end_station_name)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  theme_minimal() +
  labs(title = "Top End Stations for Member Riders", x = "Station Name", y = "Number of Rides")
```

```{r}
# Convert top member end stations to sf object
member_end_sf <- member_peak_stations %>%
  left_join(df, by = c("end_station_name")) %>% 
  distinct(end_station_name, end_lat, end_lng) %>% 
  drop_na(end_lat, end_lng) %>% 
  st_as_sf(coords = c("end_lng", "end_lat"), crs = 4326)

# Plot map
ggplot() +
  annotation_map_tile(type = "osm", zoom = 12) +
  geom_sf(data = member_end_sf, color = "red", size = 3, alpha = 0.7) +
  labs(title = "Top End Stations for Member Riders (Map)",
       x = "Longitude", y = "Latitude") +
  theme_minimal()
```

Again, looks like 2019. Members go to The Loop, Chicago's business district.

```{r}
# Extract start stations for trips ending at top member end stations
member_start_sf <- df %>%
  filter(!is.na(start_station_name)) %>% 
  filter(end_station_name %in% member_peak_stations$end_station_name) %>%
  distinct(end_station_name, start_station_name, start_lat, start_lng) %>%
  drop_na(start_lat, start_lng) %>%
  st_as_sf(coords = c("start_lng", "start_lat"), crs = 4326)

# Plot map
ggplot() +
  annotation_map_tile(type = "osm", zoom = 12) +
  geom_sf(data = member_start_sf, color = "blue", size = 3, alpha = 0.7) +
  labs(title = "Start Stations for Trips Ending at Top End Stations",
       x = "Longitude", y = "Latitude") +
  theme_minimal()

```

